---
title: "ENVS 193DS Final"
date: "June 7, 2026" 
author: "Danielle Sur" 
format:
  pdf:
    toc: true
execute:
  message: false
  warning: false
---

Link to GitHub repo [here](https://github.com/Knowknono/ENVS-193DS_spring-2026_final).

# Set up

```{r}
#| message: false

# loaded in packages
library(tidyverse)
library(janitor)
library(here)
library(readxl)

# read in data and saved as object called `nest`
nest   <- readr::read_csv(here::here("data","nest_data_final.csv"))

# read in data and saved as object called `rxntime`
rxntime   <- readr::read_csv(here::here("data","rxntime.csv"))
```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, they used linear regression. In part 2, they used a chi-square test of independence.

## b. Suggestions for rewriting

Part 1. We found that the distance of microhabitat to water edge predicts the likelihood of usage by American Avocets.

Part 2. We found that the relationship between bird species (American Avocet, Forster’s Tern, Black-necked Stilt) and habitat type (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat) is statistically significant.

We found that the prevalence of bird species (American Avocet, Forster’s Tern, Black-necked Stilt) differ significantly by habitat type (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat).

OR

We found that different bird species (American Avocet, Forster’s Tern, Black-necked Stilt) show preference for certain habitat types (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat).

## c. Further analysis

Predation pressure can be studied as a discrete variable (count of individuals); the presence of predators could make microhabitats too risky for the birds to frequent safely. As such, the birds may choose to avoid predator territories, reducing the probability of American Avocet microhabitat use in those locations. Predator data may be acquired through methods such as tracking, video sightings/ identification, and/ or direct observation.



# Problem 2. Data analysis

## a. Clean and wrangle



## b. Response variable



## c. Purpose of study



## d. Table of models



## e. Run the models



## f. Select the best model



## g. Check the diagnostics



## h. Visualize the model predictions



## i. Write a caption for your figure



## j. Calculate model predictions



## k. Interpret your results



# Problem 3. Affective and exploratory visualizations

## a. Comparing visualizations

The exploratory visualizations I made in Homework

## b. Designing an analysis



## c. Check any assumptions and run your analysis



## d. Create a visualization

### Cleaning data

```{r}
# created a new object from rxntime
rxntime_box <- rxntime |>
  # cleaned column names
  clean_names() |>
  # selected columns of interest
  select(exercise_type, reaction_time_avg_ms) |>
  # capitalized existing abbreviations
  mutate(exercise_type = case_when(
    exercise_type == "volleyball" ~ "Volleyball",
    exercise_type == "biking"    ~ "Biking",
    exercise_type == "walking"   ~ "Walking")) |>
  # removed missing values
  drop_na()
```

### Visualization

```{r}
# base layer: ggplot
ggplot(data = rxntime_box,
       # x-axis: exercise type
       # y-axis: mean reaction time
       # color by exercise type
       mapping = aes(x = exercise_type,
                     y = reaction_time_avg_ms,
                     color = exercise_type)) +
  # first layer: boxplot
  geom_boxplot() +
  # second layer: data points and jitter
  geom_jitter(width = 0.15,
              height = 0) +
  labs(title = "Exercise Type and Reaction Time",
       x = "Exercise Type",
       y = "Mean Reaction Time (ms)") +
  # custom theme
  theme_minimal() +
  # removed legend
  theme(legend.position = "none") +
  # custom colors
  scale_color_manual(values = c("Biking" = "tomato",
                                "Volleyball" = "dodgerblue",
                                "Walking" = "green3")) +
  # removed grid and axis lines
  theme(panel.grid = element_blank(),
        axis.line = element_blank())
```

## e. Write a caption



## f. Write about your results



## g. Sharing your affective visualization

I was there :)